---
title: "Exercise Prediction Algorithms (PML Coursera) "
#author: "Hélder Mendes"
date: "08/13/2014"
output: html_document
---

### Objective

Tis work will create a model to preview the quality of barbell lifts exercises performed by 6 people. Data is obtained from  <http://groupware.les.inf.puc-rio.br/har> and contain measurements from 3 accelerometers placed on the arm, belt and forearm of each participant. The classification varies from A to E where A is the best performance. The final goal is to create a model that previews reliably 20 new observations.

Two models for predition are tested one based on Support Vector Machines and a second on Random Forest.


### Loading and cleaning data


The code for loading both traning and testing sets is presented on this section. Training is a 19662x160 dimension matrix and testing is 20x160. After cleanup, the number of features is reduced to 60 on both matrices.



```{r}

################################################################################
# Loading and Cleaning Data
#
library(caret)

setwd("L:/R/practical_ml/assignements")

training <- read.table("pml-training.csv", sep=",", header=T)
testing <- read.table("pml-testing.csv", sep=",", header=T)

# Cleaning Data NA and ""
clean_flag <- apply(training, 2, function(x) {
  sum(is.na(x)) + sum(x=="")
})

clean_training <- training[,which(clean_flag == 0)]
clean_testing <- testing[,which(clean_flag == 0)]

train_index <- createDataPartition(y = clean_training$classe, p = 0.7, list = FALSE)
training_set <- clean_training[train_index, ]
cross_validation_set <- clean_training[-train_index, ]

# Remove qualitative features (fisrt 6 columns)
remove_index <- as.integer(c(1, 2, 3, 4, 5, 6))
training_set <- training_set[, -remove_index]
testing_set <- clean_testing[, -remove_index]
```

### Support Vector Machine Model

The first model to be implemented in order to predict the new 20 occurrences from the testing set is based on SVM. Code for its contruction is presented here.


Model Construction:
```{r}

mytrControlsvm = trainControl(method = "cv", number = 6)
modelsvm <- train(training_set$classe ~ ., data = training_set, 
                    method = "svmRadial", trControl = mytrControlsvm)

modelsvm

```


Prediction and accuracy:

```{r}

predicted_svm <- predict(modelsvm, cross_validation_set)
SampleError_svm <- sum(predicted_svm == cross_validation_set$classe)/nrow(cross_validation_set)
accuracy_svm = SampleError_svm * 100
prtext <- paste("Accuracy for SVM Model =", round(accuracy_svm,2), "%")
print(prtext)

```

Confusion Matrix:

```{r}
table(predicted_svm,cross_validation_set$classe)
```


Cost function versus Accuracy for SVM Model:

```{r}
plot(modelsvm)
```


Predictions:
```{r}

# Apply SVM Model to Testing Set
answers_svm <- predict(modelsvm, testing_set)
answers_svm

```




### Random Forest Model (RF)

The second model to be implemented is based on RF. The code, accuracy and corresponding predictions is presented here.


```{r}

mytrControlrf = trainControl(method = "cv", number = 4)
modelrf <- train(training_set$classe ~ ., data = training_set, 
                  method = "rf", trControl = mytrControlrf)

modelrf


```


Prediction and accuracy:

```{r}

predicted_rf <- predict(modelrf, cross_validation_set)
SampleError_rf <- sum(predicted_rf == cross_validation_set$classe)/nrow(cross_validation_set)
accuracy_rf = SampleError_rf * 100
prtext <- paste("Accuracy for RF Model =", round(accuracy_rf,2), "%")
print(prtext)

```

Confusion Matrix:

```{r}
table(predicted_rf,cross_validation_set$classe)
```


Cost function versus Accuracy for RF Model:

```{r}
plot(modelrf)
```


Predictions:


```{r}

# Apply RF Model to Testing Set
answers_rf <- predict(modelrf, testing_set)
answers_rf

```




### Conclusion

According to the results presented in this report the accuracy determined for each model is, 92.98% for Support Vector Machine method and 99.66% for Random Forest. These results are obtained from the traning set using cross-validation. The confusion matrix confirms that the number of wrong class assignements from SVM is higher than for RF.

Considering these results, the expected clesses for the new 20 occurrences from the testing set, correspond the predictions performed with the RF method, which are:


```{r}
answers_rf
```


